Repeated loss of variation in insect ovary morphology highlights the role of developmental constraint in life-history evolution

### 2020-12-08 manuscript.Rmd
updating Rmd files with new method for calculating and reporting p-values

---
title: Repeated loss of variation in insect ovary morphology highlights the role of developmental constraint in life-history evolution
output:
  bookdown::pdf_document2:
  keep_tex: yes
  latex_engine: pdflatex
  #bookdown::html_document2: default
toc: false
bibliography: manuscript_supplement.bib
csl: nature.csl
link-citations: yes
header-includes:
 \usepackage{xr} \externaldocument{manuscript_supplement}
 \usepackage{float}
 \floatplacement{figure}{H}
---

@@

The number of offspring an organism can produce is a key component of its evolutionary fitness and life-history. Here we perform a test of the hypothesized trade off between the number and size of offspring using thousands of descriptions of the number of egg-producing compartments in the insect ovary (ovarioles), a common proxy for potential offspring number in insects. In contrast to prior claims, we find that ovariole number is not generally negatively correlated with the size of insect eggs, and we highlight several factors that may have contributed to this size-number trade off being strongly asserted in previous studies. We reconstruct the evolutionary history of the nurse cell arrangement within the ovariole, and show that the diversification of ovariole number and egg size have both been largely independent of nurse cell presence or position within the ovariole. Instead we show that ovariole number evolution has been shaped by a series of transitions between variable and invariant states, with multiple independent lineages evolving to have almost no variation in ovariole number. We highlight the implications of these invariant lineages on our understanding of the specification of ovariole number during development, as well as the importance of considering developmental processes in theories of life-history evolution.

@@

@@

The dataset of insect ovariole number is available at Dryad (doi:10.5061/dryad.59zw3r253). The code and phylogenetic trees required to reproduce all the analyses, figures, and generate the manuscript files are provided at 'https://github.com/shchurch/insect_ovariole_number_evolution_2020', commit 6cf446a.

@@

All regressions were performed over a posterior distribution of trees and accounting for phenotypic uncertainty; therefore these analyses produced a range of p-values. We considered a relationship significant when the maximum p-value observed from this range was below the threshold 0.01. We report the maximum p-value throughout this manuscript, and have included the full range in the supplementary information (see Table \ref{tab:PGLS-posterior}).

@@

![**The diversity of ovariole number across insects.** a, Schematic of a generalized insect female reproductive system, showing a pair of ovaries, each with four ovarioles. b, The range of total adult ovariole number, log~10~ scale, across nine groups of insects, arranged with random jitter on the y-axis within each group. Groups are, from top to bottom: Apterygota, Palaeoptera, Polyneoptera, Condylognatha, Psocodea, Amphiesmenoptera, Antliophora, Neuropteroidea, and Hymenoptera.](manuscript_figures/Fig_Diversity_V3-01.png)

@@

Ovariole number has been hypothesized to be negatively correlated with egg size[@montague1981reproductive;@berrigan1991allometry;@starmer2003phylogenetic]. This hypothesis is based on the predictions that (1) female reproduction is resource-limited, therefore egg size should trade off with egg number, and (2) ovariole number can serve as a proxy for egg number[@montague1981reproductive;@smith1974optimal]. Our data do not support the hypothesis of a negative relationship between egg size and ovariole number across insects (p-value `r all_table[2,]$'pval_max'`; Figs. 2a, \ref{fig:PGLS-egg-ON-genus}, Table \ref{tab:PGLS-posterior}).

Given that this prediction is often conditioned on taking into account adult body size[@stewart1991reproductive;@starmer2003phylogenetic], we combined data on ovariole number and egg size with several measurements of body size, and applied a size correction that accounts for phylogenetic relationships [@revell2009size]. Across insects we did not observe a significant relationship between ovariole number and egg size when controlling for body size (p-value `r all_table[6,]$'pval_max'`; Fig. 2b, \ref{fig:PGLS-Gilbert-name}, \ref{fig:PGLS-egg-ON-family}, Table \ref{tab:PGLS-posterior}), though we do note the presence of a non-significant negative trend, primarily driven by a few extreme values (e.g. wasps in the family Trigonalidae, which have many ovarioles and lay minute eggs, Fig. \ref{fig:PGLS-egg-ON-family}).

We assessed the relationship between these traits for four sublineages of insects and found that across Drosophilidae species, egg size is indeed strongly negatively correlated with ovariole number when accounting for body size (p-value <0.001; Fig. 2c, Table \ref{tab:PGLS-posterior}). However this relationship did not hold for any of the three other lineages we analyzed (e.g. Orthoptera p-value `r all_table[9,]$'pval_max'`; Fig. 2d, \ref{fig:PGLS-Coleoptera-Hymenoptera}, Table \ref{tab:PGLS-posterior}).

We additionally tested whether ovariole number is positively correlated with adult body size, and in contrast to previous studies[@honvek1993intraspecific], we found no correlation between ovariole number and adult body weight or length across insects (p-values `r body_all_table[3,]$'pval_max'` and `r body_all_table[1,]$'pval_max'`, respectively; Fig. \ref{fig:PGLS-body-plot}, Table \ref{tab:PGLS-body-posterior}).

![**Tests of the hypothesized trade off between egg size and ovariole number.** a, Egg volume (mm^3^) and ovariole number, both log~10~ scale, n=`r nrow(egg_ON_name_pgls)` species. See section _Modeling ovariole number evolution_ for discussion of the enrichment of certain low values of ovariole number (points appearing vertically arranged) b, Egg volume and ovariole number, residuals to adult dry body weight, n=`r length(which(egg_ON_gilbert_genus_pgls$rank %in% tree$tip.label))` genera. c, Drosophilidae egg volume and ovariole number, residuals to thorax length, n=`r length(which(DIP_pgls$rank %in% dros_tree$tip.label))` species. d, Orthoptera egg volume and ovariole number, residuals to body length, n=`r length(which(ORT_pgls$rank %in% tree$tip.label))` genera.](manuscript_figures/Fig_Egg_V2-01.png)

@@

Using a previously reported dataset of lifetime fecundity measurements across insects[@Gilbert2007], we assessed the relationship between lifetime fecundity and ovariole number, and observed a positive, but not statistically significant, correlation (p-value `r all_table[8,]$'pval_max'`; Figs. 3, \ref{fig:PGLS-fecundity-name}). We interpret our results, in conjunction with those previously reported, to suggest that ovariole number, when variable across insects in a lineage, may be one factor among many influencing the number of eggs produced. However, we caution against using ovariole number as a direct mathematical proxy for offspring number.

![**The relationship between lifetime fecundity and ovariole number.** Both values are shown on a log~10~ scale. n=`r length(na.omit(fecundity_genus_pgls$fecundity))` genera, colored according to the groups shown in Fig. 1b.](manuscript_figures/Fig_Fecundity_V1-01.png)

@@

A frequently invoked life-history prediction is that, given a finite set of metabolic resources, organisms can either produce few offspring, each with high fitness, or many low-fitness offspring[@smith1974optimal;@lack1947significance;@stearns1992evolution]. In insects, egg size and ovariole number are often used as proxies for offspring fitness[@koch2014mother] and number[@david1970nombre;@bouletreau1982reproductive], respectively, and therefore it has been predicted that insects with more ovarioles lay smaller eggs than insects with fewer ovarioles[@hodin2009she;@montague1981reproductive;@berrigan1991allometry;@starmer2003phylogenetic]. Our results, using a dataset that spans `r nrow(raw_ON)` observations and that takes into account phylogenetic relationships, indicate that a general trade-off between insect egg size and ovariole number does not exist (Fig. 2).

>>>>

---
title: Repeated loss of variation in insect ovary morphology highlights the role of developmental constraint in life-history evolution
output:
  bookdown::pdf_document2:
  keep_tex: yes
  latex_engine: pdflatex
  #bookdown::html_document2: default
toc: false
bibliography: manuscript_supplement.bib
csl: nature.csl
link-citations: yes
header-includes:
 \usepackage{xr} \externaldocument{manuscript_supplement}
 \usepackage{float}
 \usepackage{array}
 \usepackage{xcolor,colortbl}
 \floatplacement{figure}{H}
---

@@

The number of offspring an organism can produce is a key component of its evolutionary fitness and life-history. Here we perform a test of the hypothesized trade off between the number and size of offspring using thousands of descriptions of the number of egg-producing compartments in the insect ovary (ovarioles), a common proxy for potential offspring number in insects. \textcolor{red}{We find evidence of a negative relationship between egg size and ovariole number when accounting for adult body size, but in contrast to prior claims, we note that this relationship is not generalizable across insect clades, and we highlight several factors that may have contributed to this size-number trade off being stated as a general rule in previous studies}. We reconstruct the evolutionary history of the nurse cell arrangement within the ovariole, and show that the diversification of ovariole number and egg size have both been largely independent of nurse cell presence or position within the ovariole. Instead we show that ovariole number evolution has been shaped by a series of transitions between variable and invariant states, with multiple independent lineages evolving to have almost no variation in ovariole number. We highlight the implications of these invariant lineages on our understanding of the specification of ovariole number during development, as well as the importance of considering developmental processes in theories of life-history evolution.

@@

\textcolor{red}{For two regressions comparing egg size to ovariole number while accounting for adult body size, we tested alternative hypotheses of evolution by simulating new data. We considered two hypotheses: no evolutionary correlation with ovariole number, and a strong correlation with ovariole number (slope of -1:1). For each trait we simulated 1,000 datasets using evolutionary parameters fit under a Brownian Motion model in the R packages geiger}[@revell2010phylogenetic]\textcolor{red}{, version `r packageVersion('geiger')`, and phylolm}[@tung2014linear]\textcolor{red}{, version `r packageVersion('phylolm')`.}

@@

The dataset of insect ovariole number is available at Dryad, doi:10.5061/dryad.59zw3r253.

## Code availability {-}

The code and phylogenetic trees required to reproduce all the analyses, figures, and generate the manuscript files are provided at 'https://github.com/shchurch/insect_ovariole_number_evolution_2020', commit 6cf446a.

@@

\textcolor{red}{All phylogenetic regressions were performed using both the maximum clade credibility (MCC) tree and over 1000 trees drawn from the posterior distribution. We considered a relationship significant when the MCC p-value was below the threshold 0.01. To assess the robustness of results to uncertainty, we also report the number of regressions over the posterior distribution of phylogenetic trees that gave a significant result (see Table \ref{tab:PGLS-posterior-egg-ON}).}

\textcolor{red}{For two comparisons, we validated that our tests had sufficient statistical power using the selected threshold by comparing the distribution of p-values from regressions of observed data to regressions of data simulated under alternative hypotheses. We used the comparison to results based on simulated data to evaluate the likelihood of false positives (comparing to data simulatd under no correlation) and false negatives (comparing to data simulated with strong correlation).}

@@

![**The diversity of ovariole number across insects.** a, Schematic of a generalized insect female reproductive system, showing a pair of ovaries, each with four ovarioles. b, The range of total adult ovariole number, log~10~ scale, across nine groups of insects, arranged with random jitter on the y-axis within each group. Groups are, from top to bottom: Apterygota, Palaeoptera, Polyneoptera, Condylognatha, Psocodea, Amphiesmenoptera, Antliophora, Neuropteroidea, and Hymenoptera.](manuscript_figures/Fig_Diversity_V4-01.png)

@@

Ovariole number has been hypothesized to be negatively correlated with egg size[@montague1981reproductive;@berrigan1991allometry;@starmer2003phylogenetic]. This hypothesis is based on the predictions that (1) female reproduction is resource-limited, therefore egg size should trade off with egg number, and (2) ovariole number can serve as a proxy for egg number[@montague1981reproductive;@smith1974optimal]. We did not observe a significant negative relationship when comparing egg size and ovariole number across insect species \textcolor{red}{(Figs. 2a, Table \ref{tab:PGLS-posterior-egg-ON}, p-value `r round(pgls_results_egg_ON_name$tTable[[8]],3)`, n=`r pgls_results_egg_ON_name$dims$N`). To test results with a larger sample size, we also compared across egg size and ovariole number across insect genera and did not observe significant relationship (Fig. \ref{fig:PGLS-egg-ON-genus}, p-value `r round(pgls_results_egg_ON_genus$tTable[[8]],3)`, n=`r pgls_results_egg_ON_genus$dims$N`).}

Given that this predicted relationship is often conditioned on taking into account adult body size[@stewart1991reproductive;@starmer2003phylogenetic], we combined data on ovariole number and egg size with several measurements of body size, and applied a size correction that accounts for phylogenetic relationships [@revell2009size]. \textcolor{red}{When accounting for adult body weight, we observed a significant negative relationship between egg size and ovariole number (Fig. 2b, \ref{fig:PGLS-Gilbert-name}, p-value `r gsub(0,"<0.001",round(pgls_results_gilbert_genus$tTable[[8]],3))`, n=`r pgls_results_gilbert_genus$dims$N`). To evaluate the robustness of this result, we repeated the analysis 1000 times, taking into account uncertainty in both the phylogeny and trait measurements. Out of 1000 regressions, `r sum(sapply(allometry_gilbert_genus,function(x){x$tTable[[8]]}) < 0.01)` indicated a significant negative relationship (Table \ref{tab:PGLS-posterior-egg-ON}). We performed the same comparison accounting for adult body length, and likewise observed a significant negative relationship (Fig. \ref{fig:PGLS-family-length}, p-value `r gsub(0,"<0.001",round(pgls_results_body_egg_ON_family$tTable[[8]],3))`, n=`r pgls_results_body_egg_ON_family$dims$N`), supported by `r sum(sapply(allometry_body_egg_ON_family,function(x){x$tTable[[8]]}) < 0.01)` of 1000 repeated analyses (Table \ref{tab:PGLS-posterior-egg-ON}).}

\textcolor{red}{We further explored these results using two methods: First, to evaluate our findings against alternative evolutionary hypotheses, we compared these results to regressions based on simulated data. Our results showed that when considering body size, the slope of the regression of egg size and ovariole number is smaller than we would observe by chance, compared to data simulated with no evolutionary correlation (Fig. \ref{fig:simulated}). When considering adult body length, but not weight, the slope of the regressions on observed data are within the range that would be expected under an strong negative relationship (-1:1 slope in log-log space, Fig. \ref{fig:simulated}). This suggests that, for at least some measures of body size, there is a strong negative relationship bewteeen egg size and ovariole number when accounting for body size.}

\textcolor{red}{Second, we assessed the relationship between egg size and ovariole number, accounting for body size, within four subclades of insects. We found that across Drosophilidae fly species, egg size is indeed strongly negatively correlated with ovariole number when accounting for body size (Fig. 2c, Table \ref{tab:PGLS-posterior-clade}, p-value `r gsub(0,"<0.001",round(pgls_results_DIP$tTable[[8]],3))`, n=`r pgls_results_DIP$dims$N`). For beetles (Coleoptera), we observed a significant relationship (\ref{fig:PGLS-Coleoptera-Hymenoptera}, p-value `r round(pgls_results_COL1$tTable[[8]],3)`, n=`r pgls_results_COL1$dims$N`), but a substantial fraction of repeat analyses show that these results are not robust to uncertainty (`r sum(sapply(allometry_COL1,function(x){x$tTable[[8]]}) > 0.01)` of 1000 regressions are not significant, Table \ref{tab:PGLS-posterior-clade}). For grasshoppers and crickets (Orthoptera) and wasps (Hymenoptera), we observed no significant relationship between ovariole number and egg size, accounting for body size (Fig. 2d, Table \ref{tab:PGLS-posterior-clade}, \ref{fig:PGLS-Coleoptera-Hymenoptera}, Orthoptera: p-value `r round(pgls_results_ORT$tTable[[8]],3)`, n=`r pgls_results_ORT$dims$N`, Hymenoptera: p-value `r round(pgls_results_HYM$tTable[[8]],3)`, n=`r round(pgls_results_HYM$dims$N)`). This indicates that, while a strong negative correlation between egg size and ovariole number exists for some insects, it does not represent a universal pattern across insect clades.}

Finally, we tested whether ovariole number is positively correlated with adult body size, and in contrast to previous studies[@honvek1993intraspecific], we found no correlation between ovariole number and adult body weight or length across insects \textcolor{red}{(Fig. \ref{fig:PGLS-body-plot}, Table \ref{tab:PGLS-posterior-body}, body weight: p-value `r round(pgls_results_body_gilbert_name$tTable[[8]],3)`, n=`r pgls_results_body_gilbert_genus$dims$N`, body length: p-value `r round(pgls_results_body_ON_family$tTable[[8]],3)`, n=`r pgls_results_body_ON_family$dims$N`). Of the four subclades considered, only insects in the family Drosophilidae and the order Orthoptera indicated a positive relationship between body size and ovariole number (Table \ref{tab:PGLS-posterior-body}, p-value `r gsub(0,"<0.001",round(pgls_results_body_ORT$tTable[[8]],3))`, n=`r pgls_results_body_ORT$dims$N`).}

![**Tests of the hypothesized trade off between egg size and ovariole number.** a, Egg volume (mm^3^) and ovariole number, both log~10~ scale, \textcolor{red}{points represent insect species}. See section _Modeling ovariole number evolution_ for discussion of the enrichment of certain low values of ovariole number (points appearing vertically arranged) b, Egg volume and ovariole number, residuals to adult dry body weight, \textcolor{red}{points=genera}. c, Drosophilidae egg volume and ovariole number, residuals to thorax length, \textcolor{red}{points=species}. d, Orthoptera egg volume and ovariole number, residuals to body length, \textcolor{red}{points=genera}.](manuscript_figures/Fig_Egg_V2-01.png)

@@

Using a previously reported dataset of lifetime fecundity measurements across insects[@Gilbert2007], we assessed the relationship between lifetime fecundity and ovariole number, and \textcolor{red}{observed a significant positive relationship (Fig. 3, p-value `r round(pgls_results_fecundity_name$tTable[[8]],3)`, n=`r pgls_results_fecundity_name$dims$N`). However, we note that this relationship is largely defined by the absence of insects with high ovariole number and low fecundity (Fig. 3, empty bottom right corner). For insects with low ovariole number, fecundity varied over more than three orders of magnitude.} We interpret our results, in conjunction with those previously reported, to suggest that ovariole number, when variable across insects in a lineage, may be one factor among many influencing the number of eggs produced. However, we caution against using ovariole number as a direct mathematical proxy for offspring number.

![**The relationship between lifetime fecundity and ovariole number.** Both values are shown on a log~10~ scale. Points \textcolor{red}{represent insect genera} and are colored according to the groups shown in Fig. 1b.](manuscript_figures/Fig_Fecundity_V1-01.png)

@@

A frequently invoked life-history prediction is that, given a finite set of metabolic resources, organisms can either produce few offspring, each with high fitness, or many low-fitness offspring[@smith1974optimal;@lack1947significance;@stearns1992evolution]. In insects, egg size and ovariole number are often used as proxies for offspring fitness[@koch2014mother] and number[@david1970nombre;@bouletreau1982reproductive], respectively, and therefore it has been predicted that insects with more ovarioles lay smaller eggs than insects with fewer ovarioles[@hodin2009she;@montague1981reproductive;@berrigan1991allometry;@starmer2003phylogenetic]. Our results, using a dataset that spans `r nrow(raw_ON)` observations and that takes into account phylogenetic relationships, indicate that a generalized trade-off between insect egg size and ovariole number does not exist (Fig. 2).
